Dynamic capabilities and competitive advantage

### 2019-10-07 article4_tommylars.rnw
still trying to nail it - looking at different levels ov VRIO

@@

The relationship between dynamic capabilities and competitive advantage has been one of
great contention in the extant literature \citep{Peteraf2013,Ambrosini2009}. From this
debate a richer understanding of how different dynamic capabilities (or functional forms
of them) work differently to change different types of firm resources
\citep{Helfat2015a,Teece2007,Ambrosini2009a} with implications for competitive advantage
\citep{Peteraf2013} has emerged. Increasingly the relationship between (1) dynamic
capabilities, (2) firm resources and routines, and competitive advantage is seen as a
complex system with interactions between different types of each 1 and 2. Many conceptual
contributions have excellently discussed varieties or functional forms of either dynamic
capabilities \citep{Ambrosini2009a,Helfat2015ad,DiStefano2014} or underlying resources or
routines \citep{Winter2003,Eisenhardt2000}. However, very few, is any, papers attempts to
integrate varieties at multiple levels simultaneously. The bulk of contributions focus
only on heterogeneity at the dynamic capabilities level, or at

>>>>

Extant literature on DC struggles with a fundamental conception and prediction of the
theory, namely if and when DC can lead to competitive advantage. This friction has been
dubbed the "elephant in the room" (peteraf) and been tracked back to the seminal papers of
Teece and Eisenhardt, as well as their their different theoretical antecedents of
evolutionary economics (ref) and behavioral theory (ref) respectively (ref Arndt
x2). Recently, distefaneo et al propopsed a theoretical integration of these varying
conceptions of DC theory. They suggest that latent and constitutent elements of DC
represent different constructs but are integrated in that they work differently in shaping
strategic change.

@@

The relationship between dynamic capabilities and competitive advantage has been one of
great contention in the extant literature \citep{Peteraf2013,Ambrosini2009}. From this
debate a richer understanding of how different dynamic capabilities (or functional forms
of them) work differently to change different types of firm resources
\citep{Helfat2015a,Teece2007,Ambrosini2009a} with implications for competitive advantage
\citep{Peteraf2013} has emerged. Increasingly the relationship between (1) dynamic
capabilities, (2) firm resources and routines, and competitive advantage is seen as a
complex system with interactions between different types of each 1 and 2. Many conceptual
contributions have excellently discussed varieties or functional forms of either dynamic
capabilities \citep{Ambrosini2009a,Helfat2015ad,DiStefano2014} or underlying resources or
routines \citep{Winter2003,Eisenhardt2000}. However, very few, if any, papers attempt to
integrate varieties at multiple levels simultaneously. By asking if dynamic capabilities
can lead to competitive advantage and how this occurs, one is far better off asking "when"
competitive advangate is an outcome of dynamic capabiliteis and its workings.

The definition of dynamic capabilities, albeit contestst, seems to be converging on
varieties of "the ability of firms to create, modicy and extend its resource base" (ref)
\citep{Schilke2018}. Meanwhile both dynamic capabilities as the "governing entity" and
resources as "the governed entity" takes different forms in the literature (ref). Dynamic
capabilities themsleves can roughly be divided into sets of simple rules governing
undelying resources, or more complex routines developed for the same purpose
\citep{Peteraf2013}. In other words, dynamic capabilities themselves can and cannot satisfy
the VRIO condition. This contention is throughly covered in the literature (refs) and
recent contribiutions have argued for interplays between simple and complex types of
dynamic capabilities foring a more complex and initmatble dynamic system
\citep{DiStefano2014}. Moreover, granted different forms of dynamic capabilities exist
their impact on the "governed entity", the underlying firm resources, should also be
varying.

To explore this we emply two radically differnet forms of dynamic capabilities / one
geared towards transforming exitsgint resources in order to exploit them to gain
competitive advantage (refs). The other form is geraed towards seizing new opportunities in order
to explore new business opportunitieis (refs). We argue that competitive advantage can
stem from each of these, but that the mechanism varies. Dynamic capabilities geared
towards transforming for exploitation ($DC_P$) is by and of itself the result of a more
complex set of routines configured in a way that is hard to imitate because of its path
dependent and evolutionary nature. This legacy from evolutinoary economics is stressed as
an important factor for creating VRIO resources (Ref peteraf). The stream of contributions
maintaining dynamic capabilities working through the object of "changing existing
capabilities" \citep{Teece1997,Zollo2002a,Winter2003,Zahra2006,Kale2002} also maintain
that

We argue that latent and constituted dynamic capabilites behave differently and work
differnetly trhough differents modes (action or wahtever the term in DiStefano er).

{\bf LAtent = complex process routines Teece
Constitutent = simpler, ad hoc - Eisenhardt}

We explore two different modes (natures) of dynamic capabilities and how they work
differently (action) to create competitive advantage. Or is it the object of action as
either resources or opportunities?

By introducing different measures of different types of DC (latent and constiuted) as well
as two different types of action (change existing or create new) we explore the workings
of the drive train model proposed by DiStefano suggesting a fuller and more dynamic view
on how latent and consittued DC works differently.

The working of latent:
Departing from the definintion of dynamic capabilities found in Teece (ref) and subsequent
contributions in this tradition

The working of cinstuted

We identify two ditintctly different noptions of dynamic capabilities both following well
known definitions byut weith different nature (ref distefano). We show that only the
latent dc can directly and indirectly create CA,whereas the constituent element nature of
DC only works through changing lower level routines.

Different parts of the drive train works on different types of underlying routines

Looks at competing hypotheses between Eisenhardt and teexe with respect to routinization. Finds that routininzation happens at DC level less than operational level?
We argue that this depends on the purpose *action of the Dc and the composition of the
focal operational capability.

Reviewing the literature reveals that researchers have tended to identify dynamic capabilities post hoc, inferring their existence from successful organizational out- comes such as profitability and growth, as prior definitions would suggest (Table

On the other hand we have dynamic capabilites geared toward sensing and seizing new
opportunities as part of an effort of exploration. Such dynamic capabilities are described
in the literature as themselves a set of more simple routines orchestrated in a manner
such that x`

Routinization at certain level, but depends on them being outward or inward looking.

How can DC lead to competitive advangate. departing from the original definition this
happens through the modification or aquisition of firm resources. But the resopurces
themselves can and cannot be a source of competitive advantage

At a minimum, in order for something to qualify as a capability, it must work in a reliable manner’ (Helfat and Peteraf, 2003: 999). Capabilities

The bulk of contributions focus
only on heterogeneity at the dynamic capabilities level, or at

W

### 2019-10-07 article4_tommylars.tex
still trying to nail it - looking at different levels ov VRIO

The relationship between dynamic capabilities and competitive advantage has been one of
great contention in the extant literature \citep{Peteraf2013,Ambrosini2009}. From this
debate a richer understanding of how different dynamic capabilities (or functional forms
of them) work differently to change different types of firm resources
\citep{Helfat2015a,Teece2007,Ambrosini2009a} with implications for competitive advantage
\citep{Peteraf2013}.  has emerged.  work dDC offers a theory of how firms achieve
competitive advantage over time. Yet the DC theory suffers from contradictory claims about
competitive advantage as its outcome. On the one hand a stream of contributions argue that
DC will lead to competitive advantage because it operates by changing operational
capabilities in face of change and hence create strategic change (ref helfat). In this
view these changes result in complex and path dependent processes that will be
idiosyncratic to the firm and hard to imitate. Consequently, they can lead to competitive
advantage.

@@

\multicolumn{1}{!{\huxvb{0, 0, 0}{0}}l!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt}\raggedright $CA_t$\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt}\raggedleft CA_{t-1}^1\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt}\raggedleft 0.88\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt}\raggedleft 0.79\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt}\raggedleft 0.88\huxbpad{6pt}} \tabularnewline[-0.5pt]

@@

\hline \\[-2pt]
 & \multicolumn{4}{c}{{\bf Models:}} \\  \cline{2-5} \\[-3pt] 
 & \emph{Direct} & \emph{Process} & \emph{Routines} & \emph{Full}\\
 & (1) & (2) & (3) & (4)\\
\hline
 &  &  &  & \\
\(CA_{t-1} \rightarrow CA\) & 0.364 & 0.361 & 0.351 & 0.343\\
 & (7.19) & (6.49) & (6.33) & (6.27)\\[3pt]
 &  &  &  & \\
\(DC_P \rightarrow CA\) & 0.378 & 0.342 & 0.314 & 0.299\\
 & (5.23) & (4.48) & (4.08) & (3.82)\\[4pt]
\(DC_X \rightarrow CA\) & -0.005 & -0.045 & 0.002 & -0.066\\
 & (-0.07) & (-0.59) & (0.04) & (-0.98)\\[4pt]
 &  &  &  & \\
\(DC_P \rightarrow P\) &  & 0.146 &  & 0.129\\
 &  & (1.93) &  & (1.7)\\[4pt]
\(DC_P \rightarrow R\) &  &  & 0.502 & 0.496\\
 &  &  & (6.23) & (6.13)\\[4pt]
\(DC_X \rightarrow P\) &  & 0.3 &  & 0.343\\
 &  & (3.63) &  & (4.79)\\[4pt]
\(DC_X \rightarrow R\) &  &  & -0.099 & -0.091\\
 &  &  & (-1.28) & (-1.14)\\[4pt]
 &  &  &  & \\
\(P \rightarrow CA\) &  & 0.182 &  & 0.185\\
 &  & (2.53) &  & (3.14)\\[4pt]
\(R \rightarrow CA\) &  &  & 0.137 & 0.123\\
 &  &  & (2.3) & (2.15)\\[4pt]
 &  &  &  & \\
\hline \\[-3pt] 
Goodness of fit & 0.45 & 0.4 & 0.43 & 0.41\\
Number of observations & 260 & 260 & 260 & 260\\
  Firm level controls & YES & YES & YES & YES\\[3pt]
\hline
 &  &  &  & \\
\end{tabular}
\end{center}
 \label{tab:reg}
\end{table}

>>>>

DC is often critizesd for being tautological and less focus has been placed on instances
when dynamism can lead to poorer competitive outcomes. 
Structural inertia, commitment \cite{schreyogg2007}. Under what conditions are changing
capabilities counterproducttive

The relationship between dynamic capabilities and competitive advantage has been one of
great contention in the extant literature \citep{Peteraf2013,Ambrosini2009}. From this
debate a richer understanding of how different dynamic capabilities (or functional forms
of them) work differently to change different types of firm resources
\citep{Helfat2015a,Teece2007,Ambrosini2009a} with implications for competitive advantage
\citep{Peteraf2013} has emerged. Increasingly the relationship between (1) dynamic
capabilities, (2) firm resources and routines, and competitive advantage is seen as a
complex system with interactions between different types of each 1 and 2. Many conceptual
contributions have excellently discussed varieties or functional forms of either dynamic
capabilities \citep{Ambrosini2009a,Helfat2015ad,DiStefano2014} or underlying resources or
routines \citep{Winter2003,Eisenhardt2000}. However, very few, if any, papers attempt to
integrate varieties at multiple levels simultaneously. By asking if dynamic capabilities
can lead to competitive advantage and how this occurs, one is far better off asking "when"
competitive advangate is an outcome of dynamic capabiliteis and its workings.

The definition of dynamic capabilities, albeit contestst, seems to be converging on
varieties of "the ability of firms to create, modicy and extend its resource base" (ref)
\citep{Schilke2018}. Meanwhile both dynamic capabilities as the "governing entity" and
resources as "the governed entity" takes different forms in the literature (ref). Dynamic
capabilities themsleves can roughly be divided into sets of simple rules governing
undelying resources, or more complex routines developed for the same purpose
\citep{Peteraf2013}. In other words, dynamic capabilities themselves can and cannot satisfy
the VRIO condition. This contention is throughly covered in the literature (refs) and
recent contribiutions have argued for interplays between simple and complex types of
dynamic capabilities foring a more complex and initmatble dynamic system
\citep{DiStefano2014}. Moreover, granted different forms of dynamic capabilities exist
their impact on the "governed entity", the underlying firm resources, should also be
varying.

To explore this we emply two radically differnet forms of dynamic capabilities / one
geared towards transforming exitsgint resources in order to exploit them to gain
competitive advantage (refs). The other form is geraed towards seizing new opportunities in order
to explore new business opportunitieis (refs). We argue that competitive advantage can
stem from each of these, but that the mechanism varies. Dynamic capabilities geared
towards transforming for exploitation ($DC_P$) is by and of itself the result of a more
complex set of routines configured in a way that is hard to imitate because of its path
dependent and evolutionary nature. This legacy from evolutinoary economics is stressed as
an important factor for creating VRIO resources (Ref peteraf). The stream of contributions
maintaining dynamic capabilities working through the object of "changing existing
capabilities" \citep{Teece1997,Zollo2002a,Winter2003,Zahra2006,Kale2002} also maintain
that

We argue that latent and constituted dynamic capabilites behave differently and work
differnetly trhough differents modes (action or wahtever the term in DiStefano er).

Looks at competing hypotheses between Eisenhardt and teexe with respect to routinization. Finds that routininzation happens at DC level less than operational level?
We argue that this depends on the purpose *action of the Dc and the composition of the
focal operational capability.

Reviewing the literature reveals that researchers have tended to identify dynamic capabilities post hoc, inferring their existence from successful organizational out- comes such as profitability and growth, as prior definitions would suggest (Table

On the other hand we have dynamic capabilites geared toward sensing and seizing new
opportunities as part of an effort of exploration. Such dynamic capabilities are described
in the literature as themselves a set of more simple routines orchestrated in a manner
such that x

Routinization at certain level, but depends on them being outward or inward looking.

How can DC lead to competitive advangate. departing from the original definition this
happens through the modification or aquisition of firm resources. But the resopurces
themselves can and cannot be a source of competitive advantage

At a minimum, in order for something to qualify as a capability, it must work in a reliable manner’ (Helfat and Peteraf, 2003: 999). Capabilities

The bulk of contributions focus
only on heterogeneity at the dynamic capabilities level, or at

W

work dDC offers a theory of how firms achieve
competitive advantage over time. Yet the DC theory suffers from contradictory claims about
competitive advantage as its outcome. On the one hand a stream of contributions argue that
DC will lead to competitive advantage because it operates by changing operational
capabilities in face of change and hence create strategic change (ref helfat). In this
view these changes result in complex and path dependent processes that will be
idiosyncratic to the firm and hard to imitate. Consequently, they can lead to competitive
advantage.

@@

\multicolumn{1}{!{\huxvb{0, 0, 0}{0}}l!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt}\raggedright $CA_{t-1}$\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt}\raggedleft CA_{t-1}^1\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt}\raggedleft 0.88\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt}\raggedleft 0.79\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt}\raggedleft 0.88\huxbpad{6pt}} \tabularnewline[-0.5pt]

@@

\hline \\[-2pt]
 & \multicolumn{4}{c}{{\bf Models:}} \\  \cline{2-5} \\[-3pt] 
 & \emph{Direct} & \emph{Process} & \emph{Routines} & \emph{Full}\\
 & (1) & (2) & (3) & (4)\\
\hline
 &  &  &  & \\
\(CA_{t-1} \rightarrow CA\) & NA & NA & NA & 0.343\\
 & (NA) & (NA) & (NA) & (6.52)\\[3pt]
 &  &  &  & \\
\(DC_P \rightarrow CA\) & 0.378 & 0.342 & 0.342 & 0.299\\
 & (5.3) & (4.78) & (4.69) & (3.77)\\[4pt]
\(DC_X \rightarrow CA\) & -0.005 & -0.045 & -0.045 & -0.066\\
 & (-0.08) & (-0.59) & (-0.6) & (-0.89)\\[4pt]
 &  &  &  & \\
\(DC_P \rightarrow P\) &  & NA &  & 0.496\\
 &  & (NA) &  & (6.07)\\[4pt]
\(DC_P \rightarrow R\) &  &  & 0.146 & 0.129\\
 &  &  & (1.97) & (1.63)\\[4pt]
\(DC_X \rightarrow P\) &  & NA &  & -0.091\\
 &  & (NA) &  & (-1.26)\\[4pt]
\(DC_X \rightarrow R\) &  &  & 0.3 & 0.343\\
 &  &  & (3.62) & (4.62)\\[4pt]
 &  &  &  & \\
\(P \rightarrow CA\) &  & NA &  & 0.123\\
 &  & (NA) &  & (2.02)\\[4pt]
\(R \rightarrow CA\) &  &  & 0.182 & 0.185\\
 &  &  & (2.72) & (2.77)\\[4pt]
 &  &  &  & \\
\hline \\[-3pt] 
Goodness of fit & 0.45 & 0.4 & 0.42 & 0.41\\
Number of observations & 260 & 260 & 260 & 260\\
  Firm level controls & YES & YES & YES & YES\\[3pt]
\hline
 &  &  &  & \\
\end{tabular}
\end{center}
 \label{tab:reg}
\end{table}
